---
title: "Bayesian SIR Model"
author: "Luis Martinez Lomeli"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

The Susceptible-Infectious-Recovered model is one of the most well known models in epidemiology and mathematical biology.
It describes the three different stages or compartments that individuals in a population can go through when they are exposed to an infectious agent. 



The classic formulation of the SIR model is based on a differential equations model written as

$\frac{dS}{dt}=-\beta S I^\alpha$

$\frac{dI}{dt}=\beta S I^\alpha - \gamma I$

$\frac{dR}{dt}=\gamma I$

$S(0)=S_0, I(0)=I_0, R(0)=R_0$

where $alpha$, $\beta$ and $\gamma$ represent the mixing, infection and recovery  rates respectively. Note that the individuals in the Recovered compartment can also be considered to be Removed as in multiple places of the literature. 


```{r, load-libraries, warning=FALSE, message=F}
rm(list=ls())

library(ggplot2)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
require(plyr)
library(reshape2)
library(deSolve)
library(utils)

RUN_STAN = FALSE  #If FALSE, then will load previously simulated samples

```

We will use a real dataset known as the `Eyam` dataset. This dataset describe the spread of the bubonic plague in the remote village of Eyam in England from June 18th to October 20th, 1666. An intresting article about this epidemic can be found [here](https://www.bbc.com/news/uk-england-35064071) 


```{r, load-Eyam, cache=TRUE}
getwd()

download.file(url = "https://github.com/msuchard/MultiBD/raw/master/data/Eyam.rda",
              destfile = "data/Eyam.rda")
load('data/Eyam.rda')
print(Eyam)

TOT_POP_EYAM = sum(Eyam[1,2:4])

df_Eyam = melt(data = Eyam, id.vars='time', variable.name = "Pop")
head(df_Eyam)

```

```{r, Eyam-plot, cache=T}

ggplot() +
  geom_point(data=df_Eyam,
           aes(x=time, y=value, group=Pop, colour=Pop) , size=2)  +
  geom_line(data=df_Eyam, aes(x=time, y=value, group=Pop, colour=Pop)) +
  theme_bw() -> lm_plot_base
print(lm_plot_base)


```


#SIR model


The first step in our analysis is to forward run the model with fixed parameters and initial conditions.

## Foward run with fixed parameters


```{r, SIR-forward-run, cache=TRUE}

# We choose population sizes and parameters values arbitrarily for illustration purposes

state = c(S = 9999, I = 1, R = 0)
TOT_POP = sum(state)
parms = c(alpha = 1, beta = 5/TOT_POP, gamma = 1)

SIR_ODE = function(t, state, parms) {
  with(as.list(c(state, parms)),{
  
  # rates of change
  dS = - beta*S*(I^alpha)
  dI = beta*S*(I^alpha) - gamma*I
  dR = gamma*I
  
  # return the rates of change
  list(c(dS, dI, dR))
}) 
}

#Solution times
times <- seq(0, 10, by = 0.25)

SIR_sol = data.frame(ode(y = state, times = times, func = SIR_ODE, parms = parms))
head(SIR_sol)


df_ode_sol = melt(data = SIR_sol, id.vars='time', variable.name = "Pop")
head(df_ode_sol)


```

```{r, SIR-forward-run-plot, cache=T}

ggplot() +
  geom_point(data=df_ode_sol,
           aes(x=time, y=value, group=Pop, colour=Pop) , size=2)  +
  
  geom_line(data=df_ode_sol, aes(x=time, y=value, group=Pop, colour=Pop)) +
  theme_bw() 

```

# Fitting the SIR model in STAN

Now, we fit the model where we especify all the elements to pass to STAN.

```{r, SIR-STAN-w-alpha, cache=TRUE}
stan_data = list(TOT_POP = TOT_POP_EYAM, TOT_SUBPOP = 3, t0 = Eyam$time[1], 
                ts = Eyam$time[-1], N_ts = length(Eyam$time[-1]), 
                y_init = as.numeric(Eyam[1, c('S', 'I', 'R')]),
                y_obs = Eyam[-1, c('S', 'I', 'R')], eps = rep(1e-5,3)
                )

cat(readLines('SIR_w-alpha.stan'), sep = '\n')

if (isTRUE(RUN_STAN)){

  stan_sir_w_alpha = stan(file = 'SIR_w-alpha.stan', data=stan_data, 
                        iter=5000, chains = 4, seed=1234)
  save(stan_sir_w_alpha, file='data/Eyam_STAN.rda.gzip', compress=TRUE)
  
} else{ 
  load(file='data/Eyam_STAN.rda.gzip')
}


traceplot(stan_sir_w_alpha, pars=c('theta_raw', 'sigma'), labels=c('alpha', 'beta', 'gamma', 'sigma'))

pairs(stan_sir_w_alpha, pars=c('theta_raw', 'sigma'), labels=c('alpha', 'beta', 'gamma', 'sigma'))

# Print information for all the parameters
# print(stan_sir_w_alpha)
# print(Eyam)

```



```{r, summary, cache=T}
print(stan_sir_w_alpha, pars=c('theta_raw', 'sigma'))
```




```{r, SIR-STAN-plot, cache=TRUE}

samples = extract(stan_sir_w_alpha)

# Create data frame of ODE solutions adding initial condition at the begining of the data frame
y_hat_summ = summary(stan_sir_w_alpha, pars=c('y_hat'))$summary
mu_summ = summary(stan_sir_w_alpha, pars=c('mu'))$summary

df_y_hat = data.frame(median=as.numeric(c(mu_summ[,"50%"],y_hat_summ[,"50%"])),
                      time = rep(Eyam$time, each=3),
                      Pop = rep(c('S', 'I', 'R'), 8)
                      )
head(df_y_hat)

# Create posterior checks summary

y_checks_summ = summary(stan_sir_w_alpha, pars=c('y_checks'))$summary

df_pred = data.frame(time=rep(Eyam$time, each=3),
                     ymin = as.numeric(y_checks_summ[,'2.5%']),
                     ymax = as.numeric(y_checks_summ[,'97.5%']),
                     ymedian = as.numeric(y_checks_summ[,'50%']),
                     ymean = as.numeric(y_checks_summ[,'mean']),
                     Pop = rep(c('S', 'I', 'R'), 8)
                    )
head(df_pred)

TOT_SAMPLES = dim(samples$y_pred_mean)[1]

# Eyam data

ggplot() +
  geom_point(data=df_Eyam,
           aes(x=time, y=value, group=Pop, colour=Pop), size=4, shape=4, stroke=2) +
  theme_bw() -> sir_plot_base
print(sir_plot_base)

# 95% Bayesian credible error bars for posterior checks

sir_plot = sir_plot_base +  
  geom_errorbar(data=df_pred, 
                aes(x=time, ymin=ymin, ymax=ymax, group=Pop, colour=Pop)) +
  ggtitle('95% Posterior Predictive Checks')
print(sir_plot)

# Fitted line
sir_plot = sir_plot_base +  
  geom_line(data = df_pred, aes(x=time, y=ymedian, group=Pop, colour=Pop), size=1,
            alpha=1) +
  geom_ribbon(data=df_pred, aes(x=time, ymin=ymin, ymax=ymax, group=Pop, fill=Pop),
              alpha=0.1,show.legend = FALSE)+
  ggtitle('95% Bayesian Credible Intervals')+
  scale_fill_manual(values=c("green", 'blue', 'red')) 
print(sir_plot)

```


## Exercise: Fit the SIR model without the mixing parameter alpha

```{r, SIR-STAN, cache=TRUE, include=F}



# stan_sir = stan(file = 'SIR.stan', data=stan_data, iter=1500, chains = 4, seed=1234)

# stan_sir = stan(model_code=stan_SIR, data=stan_data, iter=5000)
# 
# print(stan_sir, pars=c('theta', 'mu', 'sigma'))
# 
# pairs(stan_sir, pars=c('theta', 'sigma'))
# 
# print(stan_sir, pars=c('y_hat'))
# 
# print(stan_sir)
# 
# print(Eyam)

```








# Session Info
```{r}
library(utils)
sessionInfo()
```

